\title{Rarities in phonetics and phonology}
\subtitle{Structural, typological, evolutionary, and social dimensions}
\BackBody{Rare phenomena play a key role in forming and challenging linguistic theory. This volume presents multi-faceted analyses of rarities in phonetics and phonology, from a wide variety of theoretical standpoints. Some contributions to the volume analyse language-specific rare features, placing them in a broader cross-linguistic context and looking at a sum of their phonological, phonetic, and evolutionary properties, at times also making connections to sociolinguistic factors. Others consider the same (or similar) phenomena from different analytical angles, with extensive cross-referencing, or take a broad analytical or typological stance towards rare phenomena and discuss what it means to be rare.

The volume provides a nuanced picture of phonetic and phonological rarities in genealogically diverse languages, mostly lesser-studied, from around the globe. Authors were encouraged to attempt to strike a middle ground between radical exoticisation of the rarities at hand (describing them in idiosyncratic terms) and radical normalisation (underplaying the rarity of the phenomena at hand). Highly theory-specific or technical terminology is avoided or explained carefully, in order to make the book maximally accessible for a wide typologically-minded audience.
}
\author{Natalia Kuznetsova and Cormac Anderson and Shelece Easterday} 

\renewcommand{\lsImpressumExtra}{\vskip\smallskipamount
    Universita' Cattolica del Sacro Cuore contributed in 2024 to the funding of this research project and its publication.
    \smallskip\\}

\renewcommand{\lsISBNdigital}{978-3-96110-498-7}
\renewcommand{\lsISBNhardcover}{978-3-98554-127-0}
\BookDOI{10.5281/zenodo.14712629}
\typesetter{Cormac Anderson, Sebastian Nordhoff}
\proofreader{Amir Ghorbanpour,
Christopher Straughn,
Craevschi Alexandru,
Elliott Pearl,
Felix Kopecky,
Jean Nitzke,
Jeroen van de Weijer,
Mary Ann Walter,
Michael Rießler,
Neneng Sri Wahyuningsih,
Olga Olina,
Sebastian Nordhoff,
Silvie Strauß,
Tihomir Rangelov,
Timofey Mets,
Wilson Lui,
Yvonne Treis
}


\renewcommand{\lsID}{415}

\renewcommand{\lsSeries}{tpd} 
\renewcommand{\lsSeriesNumber}{5}
